---
title: "Triple Crown Motocross - 2024 Report"
format: 
  html:
    toc: true
---

# 1 Abstract

The Triple Crown Series is Canada's premier off-road motorcycle racing series. Data including the places and lap times of riders are freely available online, provided by [Trackside](https://www.tracksidesoftware.com/). However, there exists no published analysis on the series. This report aims to take these first steps and do some exploratory data analysis.

# 2 Background information

The Triple Crown Series is named in reference to the three main events in the sport of off-road motorcycle racing. There is motocross (outdoors, usually featuring longer courses), supercross (usually compact, shorter courses), and arenacross (supercross but restricted to an arena or stadium). Riders take part in rounds of these three types of races throughout the year, accumulating points in attempt to win the championship. In 2024, only motocross events took place.

There are three main classes for the Triple Crown Series. There is one class for women, WMX. There are two classes for the men, 250 Pro/Am and 450 Pro. The 250 class of riders are a mixture of amateur and professional riders, racing 250cc dirt bikes. The 450 class of riders are professional riders, racing 450cc dirt bikes.

# 3 Data scraping and tidying

Importing relevant packages:

```{r}
#| output: false
library(tidyverse) # data cleaning
library(rvest) # data scraping
library(stringr) # string operations
library(lubridate) # date/time operations
library(gt) # tables
library(ggplot2) # visuals
library(expm) # matrix operations
library(purrr)
```

Importing rider information spreadsheet:

```{r}
riders <- read.csv("rider_info.csv")
glimpse(riders)
```

The individual moto results table does not display any information about points scored. The following is a self-created tibble that will eventually get added. This will allow for analysis about points scored throughout a season.

```{r}
points <- tibble(finish = 1:40, 
                 points = c(25,22,20,18,16,15,14,13,12,11,10,9,8,7,6,5,4,3,2,1,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0))

points |>
  slice_head(n = 20) |>
  gt() |>
  cols_label(finish = "Finish",
             points = "Points")
```

Here is a function that converts the lap times (string format) to seconds (numeric format). For example, a lap time of 1:13.21 gets converted to 73.21.

```{r}
convert_to_seconds <- function(time_string) {
  if (is.na(time_string) | time_string == "") {
    return(NA)
  }
  
  if (grepl(":", time_string)) {
    mins <- as.numeric(substr(time_string, 1, 
                              regexpr(":", time_string) - 1))
    secs <- as.numeric(substr(time_string, 
                              regexpr(":", time_string) + 1,
                              nchar(time_string)))
    
    return(mins * 60 + secs)
  } 
  
  else {
    numeric_val <- as.numeric(time_string)
    
    return(numeric_val)
  }
}
```

Here is the function to scrap the data from the Trackside tables.

```{r}
mx_scraping <- function(url, year, round, moto, track,
                        date, raceid, class, type) {
  webpage <- read_html(url)
  lap_times_table <- webpage |>
    html_nodes("table") |>
    html_table(fill = TRUE)

  lap_times <- lap_times_table[[8]]

  lap_times <- lap_times |>
    as_tibble() |>
    slice(3:n())

  lap <- paste0("L", 1:(ncol(lap_times)-1))
  lap2 <- append("name", lap)
  lap_times_reduced <- lap_times |>
    setNames(lap2)

  check <- lap_times_reduced$name[1]
  if (str_detect(check, "\r\n\t\t\t\t\t\t\r\n\t\t\t\t\t\t")) {
      lap_times_reduced <- lap_times_reduced |>
        separate_wider_delim(
        name,
        delim = "\r\n\t\t\t\t\t\t\r\n\t\t\t\t\t\t",
        names = c("name1", "name2"),
        too_few = "align_start"
      ) |>
      separate_wider_delim(
        name1,
        delim = " - ",
        names = c("finish", "rider"),
        too_few = "align_start"
      ) |>
      separate_wider_delim(
        name2,
        delim = "\t\r\n\t\t\t\t\t ",
        names = c("location", "number"),
        too_few = "align_start"
      ) |>
      drop_na(rider) |>
      mutate(
        finish = seq_len(nrow(lap_times_reduced)-1),
        number = str_remove(number, "#")
      )
  } else {
  lap_times_reduced <- lap_times_reduced |>
    separate_wider_delim(
      name,
      delim = " - ",
      names = c("finish", "name"),
      too_few = "align_start"
      ) |>
    separate_wider_delim(
      name,
      delim = "#",
      names = c("rider", "number")
    ) |>
    drop_na(rider) |>
    mutate(rider = str_trim(rider),
           finish = seq_len(nrow(lap_times_reduced)-1),
           location = NA
           )
  }

  lap_times_reduced2 <- lap_times_reduced
  for (col in lap) {
    lap_times_reduced2 <- lap_times_reduced2 |>
      separate_wider_delim(
        col,
        delim = "\r\n\t\t\t\t\t",
        names = c(paste0(col, "_time"), paste0(col, "_behind"), 
                paste0(col, "_place")),
        too_few = "align_start"
      )
  }

  lap3 <- paste0("L", 1:length(lap), "_place")
  lap_times_reduced2 <- lap_times_reduced2 |>
    mutate(across(lap3, ~ str_extract(.x, "\\d+") |> 
                  as.numeric(), .names = "{col}"))

  lap4 <- paste0("L", 1:length(lap), "_time")
  lap5 <- paste0("L", 1:length(lap), "_behind")
  final_lap_times <- lap_times_reduced2
  for (i in lap4) {
    final_lap_times <- final_lap_times |> 
      mutate(!!sym(i) := str_trim(!!sym(i)),
             !!sym(i) := sapply(!!sym(i), convert_to_seconds))
  }
  for (i in lap5) {
    final_lap_times <- final_lap_times |> 
      mutate(!!sym(i) := str_trim(!!sym(i)),
             !!sym(i) := sapply(!!sym(i), convert_to_seconds))
  }

  final_lap_times <- final_lap_times |>
    pivot_longer(
      cols = starts_with("L"), 
      names_to = c("lap", ".value"), 
      names_pattern = "L(\\d+)_(time|behind|place)",
      names_repair = "unique"
    ) |>
    mutate(lap = as.numeric(lap),
           year = year,
           round = round,
           moto = moto,
           track = track,
           date = parse_date(date, "%m/%d/%Y"),
           race_id = raceid,
           class = class,
           type = type) |>
    left_join(points) |>
    drop_na(time)
  
  return(final_lap_times)

}
```

From here, we can apply this function for every moto of each class.

<u><u>450 class</u></u>:

```{r}
#| warning: false
rd11_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14641&rn=1&rt=M", 2024, 1, 1, "Wild Rose MX", "06/02/2024", "450_2024_1", "450", "mx")

rd12_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14641&rn=2&rt=M", 2024, 1, 2, "Wild Rose MX", "06/02/2024", "450_2024_2", "450", "mx")

rd21_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14683&rn=1&rt=M", 2024, 2, 1, "Lethbridge MX", "06/09/2024", "450_2024_3", "450", "mx")

rd22_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14683&rn=2&rt=M", 2024, 2, 2, "Lethbridge MX", "06/09/2024", "450_2024_4", "450", "mx")

rd31_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14767&rn=1&rt=M", 2024, 3, 1, "Prairie Hill MX", "06/16/2024", "450_2024_5", "450", "mx")

rd32_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14767&rn=2&rt=M", 2024, 3, 2, "Prairie Hill MX", "06/16/2024", "450_2024_6", "450", "mx")

rd41_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14774&rn=1&rt=M", 2024, 4, 1, "Gopher Dunes", "07/07/2024", "450_2024_7", "450", "mx")

rd42_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14774&rn=2&rt=M", 2024, 4, 2, "Gopher Dunes", "07/07/2024", "450_2024_8", "450", "mx")

rd51_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14781&rn=1&rt=M", 2024, 5, 1, "Sand Del Lee MX101", "07/14/2024", "450_2024_9", "450", "mx")

rd52_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14781&rn=2&rt=M", 2024, 5, 2, "Sand Del Lee MX101", "07/14/2024", "450_2024_10", "450", "mx")

rd61_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14788&rn=1&rt=M", 2024, 6, 1, "Riverglade MX", "07/19/2024", "450_2024_11", "450", "mx")

rd62_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14788&rn=2&rt=M", 2024, 6, 2, "Riverglade MX", "07/19/2024", "450_2024_12", "450", "mx")

rd71_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14795&rn=1&rt=M", 2024, 7, 1, "MX Deschambault", "07/26/2024", "450_2024_13", "450", "mx")

rd72_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14795&rn=2&rt=M", 2024, 7, 2, "MX Deschambault", "07/26/2024", "450_2024_14", "450", "mx")

rd81_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14802&rn=1&rt=M", 2024, 8, 1, "Walton Raceway", "08/09/2024", "450_2024_15", "450", "mx")

rd82_2024_450 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=33&e=14802&rn=2&rt=M", 2024, 8, 2, "Walton Raceway", "08/09/2024", "450_2024_16", "450", "mx")

all_2024_450 <- bind_rows(rd11_2024_450, rd12_2024_450,
                          rd21_2024_450, rd22_2024_450,
                          rd31_2024_450, rd32_2024_450,
                          rd41_2024_450, rd42_2024_450,
                          rd51_2024_450, rd52_2024_450,
                          rd61_2024_450, rd62_2024_450,
                          rd71_2024_450, rd72_2024_450,
                          rd81_2024_450, rd82_2024_450)
```

<u>250 class</u>:

```{r}
#| warning: false
rd11_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14641&rn=1&rt=M", 2024, 1, 1, "Wild Rose MX", "06/02/2024", "250_2024_1", "250", "mx")

rd12_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14641&rn=2&rt=M", 2024, 1, 2, "Wild Rose MX", "06/02/2024", "250_2024_2", "250", "mx")

rd21_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14683&rn=1&rt=M", 2024, 2, 1, "Lethbridge MX", "06/09/2024", "250_2024_3", "250", "mx")

rd22_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14683&rn=2&rt=M", 2024, 2, 2, "Lethbridge MX", "06/09/2024", "250_2024_4", "250", "mx")

rd31_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14767&rn=1&rt=M", 2024, 3, 1, "Prairie Hill MX", "06/16/2024", "250_2024_5", "250", "mx")

rd32_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14767&rn=2&rt=M", 2024, 3, 2, "Prairie Hill MX", "06/16/2024", "250_2024_6", "250", "mx")

rd41_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14774&rn=1&rt=M", 2024, 4, 1, "Gopher Dunes", "07/07/2024", "250_2024_7", "250", "mx")

rd42_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14774&rn=2&rt=M", 2024, 4, 2, "Gopher Dunes", "07/07/2024", "250_2024_8", "250", "mx")

rd51_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14781&rn=1&rt=M", 2024, 5, 1, "Sand Del Lee MX101", "07/14/2024", "250_2024_9", "250", "mx")

rd52_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14781&rn=2&rt=M", 2024, 5, 2, "Sand Del Lee MX101", "07/14/2024", "250_2024_10", "250", "mx")

rd61_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14788&rn=1&rt=M", 2024, 6, 1, "Riverglade MX", "07/19/2024", "250_2024_11", "250", "mx")

rd62_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14788&rn=2&rt=M", 2024, 6, 2, "Riverglade MX", "07/19/2024", "250_2024_12", "250", "mx")

rd71_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14795&rn=1&rt=M", 2024, 7, 1, "MX Deschambault", "07/26/2024", "250_2024_13", "250", "mx")

rd72_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14795&rn=2&rt=M", 2024, 7, 2, "MX Deschambault", "07/26/2024", "250_2024_14", "250", "mx")

rd81_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14802&rn=1&rt=M", 2024, 8, 1, "Walton Raceway", "08/09/2024", "250_2024_15", "250", "mx")

rd82_2024_250 <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=32&e=14802&rn=2&rt=M", 2024, 8, 2, "Walton Raceway", "08/09/2024", "250_2024_16", "250", "mx")

all_2024_250 <- bind_rows(rd11_2024_250, rd12_2024_250,
                          rd21_2024_250, rd22_2024_250,
                          rd31_2024_250, rd32_2024_250,
                          rd41_2024_250, rd42_2024_250,
                          rd51_2024_250, rd52_2024_250,
                          rd61_2024_250, rd62_2024_250,
                          rd71_2024_250, rd72_2024_250,
                          rd81_2024_250, rd82_2024_250)
```

<u>WMX class</u>:

```{r}
#| warning: false
rd11_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=54&e=14641&rn=1&rt=M", 2024, 1, 1, "Wild Rose MX", "06/02/2024", "wmx_2024_1", "WMX", "mx")

rd12_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=54&e=14641&rn=2&rt=M", 2024, 1, 2, "Wild Rose MX", "06/02/2024", "wmx_2024_2", "WMX", "mx")

rd21_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=54&e=14683&rn=1&rt=M", 2024, 2, 1, "Lethbridge MX", "06/09/2024", "wmx_2024_3", "WMX", "mx")

rd22_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=54&e=14683&rn=2&rt=M", 2024, 2, 2, "Lethbridge MX", "06/09/2024", "wmx_2024_4", "WMX", "mx")

rd31_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=54&e=14767&rn=1&rt=M", 2024, 3, 1, "Prairie Hill MX", "06/16/2024", "wmx_2024_5", "WMX", "mx")

rd32_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=54&e=14767&rn=2&rt=M", 2024, 3, 2, "Prairie Hill MX", "06/16/2024", "wmx_2024_6", "WMX", "mx")

# Note that there are no WMX lap times for at Gopher Dunes so only scraping the results

url <- "https://resultsmx.com/cmrc-archive-2024/class.asp?c=58&e=14774"
webpage <- read_html(url)
lap_times_table <- webpage |>
  html_nodes("table") |>
  html_table(fill = TRUE)

lap_times <- lap_times_table[[8]]

rd4_2024_wmx <- lap_times |>
  as_tibble() |>
  slice(2:n()) |>
  separate(X1, into = c("place", "number"), sep = "\r\n\t\t\t\r\n\t\t\t\t#") |>
  select(-place, -X2, -X6) |>
  rename(rider = X3,
         moto1_finish = X4,
         moto2_finish = X5) |>
  pivot_longer(
    cols = c(moto1_finish, moto2_finish), 
    names_to = "moto", 
    values_to = "finish") |>
  mutate(moto = str_extract(moto, "\\d+"),
         finish = as.numeric(gsub("(st|nd|rd|th)$", "", finish)),
         number = str_remove(number, "#")
         ) |>
  mutate(lap = NA,
         time = NA,
         behind = NA,
         place = NA,
         year = 2024,
         moto = as.numeric(moto),
         round = 4,
         track = "Gopher Dunes",
         date = parse_date("07/07/2024", "%m/%d/%Y"),
         race_id = ifelse(moto == 1, "wmx_2024_7", "wmx_2024_8"),
         class = "WMX",
         type = "mx") |>
  left_join(points) |>
  select(finish, rider, number, lap, time, behind, place,
         year, round, moto, track, date, race_id, class, type, points) |>
  mutate(rider = str_trim(rider) |> str_squish())

rd51_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=58&e=14781&rn=1&rt=M", 2024, 5, 1, "Sand Del Lee MX101", "07/14/2024", "wmx_2024_9", "WMX", "mx")

rd52_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=58&e=14781&rn=2&rt=M", 2024, 5, 2, "Sand Del Lee MX101", "07/14/2024", "wmx_2024_10", "WMX", "mx")

rd61_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=58&e=14788&rn=1&rt=M", 2024, 6, 1, "Riverglade MX", "07/19/2024", "wmx_2024_11", "WMX", "mx")

rd62_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=58&e=14788&rn=2&rt=M", 2024, 6, 2, "Riverglade MX", "07/19/2024", "wmx_2024_12", "WMX", "mx")

rd71_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=58&e=14795&rn=1&rt=M", 2024, 7, 1, "MX Deschambault", "07/26/2024", "wmx_2024_13", "WMX", "mx")

rd72_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=58&e=14795&rn=2&rt=M", 2024, 7, 2, "MX Deschambault", "07/26/2024", "wmx_2024_14", "WMX", "mx")

rd81_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=83&e=14802&rn=1&rt=M", 2024, 8, 1, "Walton Raceway", "08/09/2024", "wmx_2024_15", "WMX", "mx")

rd82_2024_wmx <- mx_scraping("https://resultsmx.com/cmrc-archive-2024/laptimes.asp?s=&c=83&e=14802&rn=2&rt=M", 2024, 8, 2, "Walton Raceway", "08/09/2024", "wmx_2024_16", "WMX", "mx")

all_2024_wmx <- bind_rows(rd11_2024_wmx, rd12_2024_wmx,
                          rd21_2024_wmx, rd22_2024_wmx,
                          rd31_2024_wmx, rd32_2024_wmx,
                          rd4_2024_wmx,
                          rd51_2024_wmx, rd52_2024_wmx,
                          rd61_2024_wmx, rd62_2024_wmx,
                          rd71_2024_wmx, rd72_2024_wmx,
                          rd81_2024_wmx, rd82_2024_wmx)
```

Merge all three classes and then writing to a csv to save.

```{r}
all_2024 <- bind_rows(all_2024_450, all_2024_250, all_2024_wmx)
write.csv(all_2024,"all_2024.csv")
```

Join in rider information for a more complete dataset.

```{r}
all_2024_cleaned <- read.csv("all_2024_cleaned.csv")

all_2024_complete <- all_2024_cleaned |>
  cross_join(riders) |>
  filter(str_detect(rider, name) & class.x == class.y &
           year.x == year.y) |>
  select(name, location, number, finish, lap, time, behind, place,
         year.x, round, moto, track, date, race_id, class.x, 
         type, points) |>
  rename(rider = name, year = year.x, class = class.x) |>
  mutate(date = ymd(date))

write.csv(all_2024_complete,"all_2024_complete.csv")

```

# 4 Analysis

## 4.1 Demographic summary

Distributions of home locations:

```{r}
all_2024_complete |>
  distinct(class, rider, location) |>
  group_by(class, location) |>
  summarize(n = n(),
            .groups = "drop") |>
  ungroup() |>
  arrange(desc(n)) |>
  pivot_wider(names_from = class,
              values_from = n) |>
  filter(! location == "unknown") |>
  mutate(`250` = if_else(is.na(`250`), 0, `250`),
         WMX = if_else(is.na(WMX), 0, WMX),
         `450` = if_else(is.na(`450`), 0, `450`)) |>
  select(location, `450`, `250`, WMX) |>
  gt() |>
  cols_label(location = "Location",
             `450` = "450",
             `250` = "250",
             WMX = "WMX")
```

## 4.2 Passing probabilities

**Q1**: What is the likelihood to go from one place to any other place on any given lap?

Let's start with the 450 class.

Formatting the empirical data:

```{r}
#| warning: false
all_2024_450_add <- all_2024_complete |>
  filter(class == "450") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  group_by(place, place_next) |>
  summarise(n = n()) |>
  ungroup()

all_2024_450_add2 <- all_2024_450_add |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_450_add <- all_2024_450_add |>
  left_join(all_2024_450_add2) |>
  mutate(prob = n/total)
```

Creating the transition matrix:

1.  Individual place positions - We have a 30 x 30 matrix since 450 motos in 2024 included a maximum of 30 riders.

```{r}
rows = 30
cols = 30
cats <- 1:30

matrix450_1 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2024_450_add |> 
      filter(place == i & place_next == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_450_add |> 
        filter(place == i & place_next == j) |> pull(prob)
    }
    matrix450_1[i, j] <- matrix_value
  }
}

rownames(matrix450_1) <- cats
colnames(matrix450_1) <- cats
print(round(matrix450_1, 2))
```

We can see that there is a 96% chance of staying in first and a 3% chance of falling from first to second on any given lap.

2.  Grouped place positions - The 30 x 30 matrix can be overwhelming. So, let's group the place positions and apply the same method. Here are the group we will use: 1-5, 6-10, 11-15, 16-20, 21-25, 26-30.

```{r}
#| warning: false
all_2024_450_placecat <- all_2024_complete |>
  filter(class == "450") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, 
                            "21-25","26-30"))))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20",
                          ifelse(place_next >= 21 & place_next <= 25, "21-25", "26-30")))))) |>
  select(place_cat, place_next_cat) |>
  group_by(place_cat, place_next_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2024_450_placecat2 <- all_2024_450_placecat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_450_placecat <- all_2024_450_placecat |>
  left_join(all_2024_450_placecat2) |>
  mutate(prob = n/total)

rows2 = 6
cols2 = 6
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")

matrix450_2 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2024_450_placecat |>
      filter(place_cat == cats2[i] & place_next_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_450_placecat |> 
        filter(place_cat == cats2[i] & place_next_cat == cats2[j]) |> pull(prob)
    }
    matrix450_2[i, j] <- matrix_value
  }
}

rownames(matrix450_2) <- cats2
colnames(matrix450_2) <- cats2

tm_450 <- as.data.frame(as.table(matrix450_2))
ggplot(tm_450, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Passing Probabilities",
       subtitle = "2024 Season - 450 class",
       x = "From Place", 
       y = "To Place", fill = "Probability")
```

Let's apply the same analysis to the 250 class. The only difference will be that 250 motos can include a maximum of 40 riders so the groups will be 1-5, 6-10, 11-15, 16-20, 21-25, 26-30, 31-35, 36-40.

```{r}
#| warning: false
all_2024_250_add <- all_2024_complete |>
  filter(class == "250") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  group_by(place, place_next) |>
  summarise(n = n()) |>
  ungroup()

all_2024_250_add2 <- all_2024_250_add |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_250_add <- all_2024_250_add |>
  left_join(all_2024_250_add2) |>
  mutate(prob = n/total)

rows = 40
cols = 40
cats <- 1:40

matrix250_1 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2024_250_add |> 
      filter(place == i & place_next == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_250_add |> 
        filter(place == i & place_next == j) |> pull(prob)
    }
    matrix250_1[i, j] <- matrix_value
  }
}

rownames(matrix250_1) <- cats
colnames(matrix250_1) <- cats
print(round(matrix250_1, 2))
```

```{r}
#| warning: false
all_2024_250_placecat <- all_2024_complete |>
  filter(class == "250") |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, "21-25",
                     ifelse(place >= 26 & place <= 30, "26-30",
                     ifelse(place >= 31 & place <= 35,
                            "31-35", "36-40"))))))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20",
                          ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                          ifelse(place_next >= 26 & place_next <= 30, "26-30",
                          ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
  select(place_cat, place_next_cat) |>
  group_by(place_cat, place_next_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2024_250_placecat2 <- all_2024_250_placecat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_250_placecat <- all_2024_250_placecat |>
  left_join(all_2024_250_placecat2) |>
  mutate(prob = n/total)

rows2 = 8
cols2 = 8
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
           "31-35", "36-40")

matrix250_2 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2024_250_placecat |>
      filter(place_cat == cats2[i] & place_next_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_250_placecat |> 
        filter(place_cat == cats2[i] & place_next_cat == cats2[j]) |> pull(prob)
    }
    matrix250_2[i, j] <- matrix_value
  }
}

rownames(matrix250_2) <- cats2
colnames(matrix250_2) <- cats2

tm_250 <- as.data.frame(as.table(matrix250_2))
ggplot(tm_250, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Passing Probabilities",
       subtitle = "2024 Season - 250 class",
       x = "From Place", 
       y = "To Place", fill = "Probability")
```

Lastly, let's apply the same analysis to the WMX class.

```{r}
#| warning: false
all_2024_wmx_add <- all_2024_complete |>
  filter(class == "WMX") |>
  drop_na(place) |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  group_by(place, place_next) |>
  summarise(n = n()) |>
  ungroup()

all_2024_wmx_add2 <- all_2024_wmx_add |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_wmx_add <- all_2024_wmx_add |>
  left_join(all_2024_wmx_add2) |>
  mutate(prob = n/total)

rows = 30
cols = 30
cats <- 1:30

matrixwmx_1 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2024_wmx_add |> 
      filter(place == i & place_next == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_wmx_add |> 
        filter(place == i & place_next == j) |> pull(prob)
    }
    matrixwmx_1[i, j] <- matrix_value
  }
}

rownames(matrixwmx_1) <- cats
colnames(matrixwmx_1) <- cats
print(round(matrixwmx_1, 2))
```

```{r}
#| warning: false
all_2024_wmx_placecat <- all_2024_complete |>
  filter(class == "WMX") |>
  drop_na(place) |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  select(place, place_next) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, 
                            "21-25","26-30"))))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20",
                          ifelse(place_next >= 21 & place_next <= 25, "21-25", "26-30")))))) |>
  select(place_cat, place_next_cat) |>
  group_by(place_cat, place_next_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2024_wmx_placecat2 <- all_2024_wmx_placecat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_wmx_placecat <- all_2024_wmx_placecat |>
  left_join(all_2024_wmx_placecat2) |>
  mutate(prob = n/total)

rows2 = 6
cols2 = 6
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")

matrixwmx_2 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2024_wmx_placecat |>
      filter(place_cat == cats2[i] & place_next_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_wmx_placecat |> 
        filter(place_cat == cats2[i] & place_next_cat == cats2[j]) |> pull(prob)
    }
    matrixwmx_2[i, j] <- matrix_value
  }
}

rownames(matrixwmx_2) <- cats2
colnames(matrixwmx_2) <- cats2

tm_wmx <- as.data.frame(as.table(matrixwmx_2))
ggplot(tm_wmx, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Passing Probabilities",
       subtitle = "2024 Season - WMX class",
       x = "From Place", 
       y = "To Place", fill = "Probability")
```

The same analysis can be done at an individual moto or individual rider level. Here are the passing probabilities for the first 450 moto of 2024.

```{r}
#| warning: false
pass_prob_moto_fn <- function(input_class, input_raceid) {
  placecat <- all_2024_complete |>
    filter(class == input_class & race_id == input_raceid) |>
    mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
    mutate(place_next = lead(place, n = 1L,
                             order_by = race_id_name)) |>
    group_by(race_id_name) |>
    mutate(place_next = c(place_next[-n()], NA)) |>
    ungroup() |>
    drop_na(place, place_next) |>
    select(place, place_next) |>
    mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                              ifelse(place >= 6 & place <= 10, "6-10",
                                     ifelse(place >= 11 & place <= 15, "11-15",
                                            ifelse(place >= 16 & place <= 20, "16-20",
                                                   ifelse(place >= 21 & place <= 25, "21-25",
                                                          ifelse(place >= 26 & place <= 30, "26-30",
                                                                 ifelse(place >= 31 & place <= 35,
                                                                        "31-35", "36-40"))))))),
           place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                                   ifelse(place_next >= 6 & place_next <= 10, "6-10",
                                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                                                 ifelse(place_next >= 16 & place_next <= 20, "16-20",
                                                        ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                                                               ifelse(place_next >= 26 & place_next <= 30, "26-30",
                                                                      ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
    select(place_cat, place_next_cat) |>
    group_by(place_cat, place_next_cat) |>
    summarise(n = n()) |>
    ungroup()
  
  placecat2 <- placecat |>
    group_by(place_cat) |>
    summarise(total = sum(n)) |>
    ungroup()
  
  placecat <- placecat |>
    left_join(placecat2) |>
    mutate(prob = n/total)
  
  distinct_cat <- placecat %>%
    summarise(distinct_cats = n_distinct(place_cat)) |>
    pull(distinct_cats)
  
    rows = distinct_cat
    cols = distinct_cat
    if (distinct_cat == 4) {
      cats <- c("1-5", "6-10", "11-15", "16-20")
    } else if (distinct_cat == 5) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25")
    } else if (distinct_cat == 6) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")
    } else if (distinct_cat == 7) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
                "31-35")
    } else if (distinct_cat == 8) {
      cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
                "31-35", "36-40")
    }
    
    pass_matrix <- matrix(0, nrow = rows, ncol = cols)
    for (i in 1:rows) {
      for (j in 1:cols) {
        matrix_row <- placecat |>
          filter(place_cat == cats[i] & place_next_cat == cats[j])
        if (nrow(matrix_row) == 0) {
          matrix_value <- 0
        } else {
          matrix_value <- placecat |> 
            filter(place_cat == cats[i] & place_next_cat == cats[j]) |> pull(prob)
        }
        pass_matrix[i, j] <- matrix_value
      }
    }
    
    rownames(pass_matrix) <- cats
    colnames(pass_matrix) <- cats

  pass_matrix_df <- as.data.frame(as.table(pass_matrix))
  ggplot(pass_matrix_df, aes(Var1, Var2, fill = Freq)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "#da3a32") +
    geom_text(aes(label = round(Freq,2))) +
    labs(title = "Passing Probabilities",
         subtitle = input_raceid,
         x = "From Place", 
         y = "To Place", fill = "Probability")
}
  

pass_prob_moto_fn("450", "450_2024_1")
```

Here are 450 rider Tanner Ward's passing probabilities across the entire 2024 season.

```{r}
#| warning: false
pass_prob_rider_fn <- function(input_rider) {
  placecat <- all_2024_complete |>
    filter(rider == input_rider) |>
    mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
    mutate(place_next = lead(place, n = 1L,
                             order_by = race_id_name)) |>
    group_by(race_id_name) |>
    mutate(place_next = c(place_next[-n()], NA)) |>
    ungroup() |>
    drop_na(place, place_next) |>
    select(place, place_next) |>
    mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                              ifelse(place >= 6 & place <= 10, "6-10",
                                     ifelse(place >= 11 & place <= 15, "11-15",
                                            ifelse(place >= 16 & place <= 20, "16-20",
                                                   ifelse(place >= 21 & place <= 25, "21-25",
                                                          ifelse(place >= 26 & place <= 30, "26-30",
                                                                 ifelse(place >= 31 & place <= 35, "31-35", "36-40"))))))),
           place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                                   ifelse(place_next >= 6 & place_next <= 10, "6-10",
                                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                                                 ifelse(place_next >= 16 & place_next <= 20, "16-20",
                                                        ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                                                               ifelse(place_next >= 26 & place_next <= 30, "26-30",
                                                                      ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
    select(place_cat, place_next_cat) |>
    group_by(place_cat, place_next_cat) |>
    summarise(n = n()) |>
    ungroup()
  
  placecat2 <- placecat |>
    group_by(place_cat) |>
    summarise(total = sum(n)) |>
    ungroup()
  
  placecat <- placecat |>
    left_join(placecat2) |>
    mutate(prob = n/total)
  
  distinct_cat <- placecat %>%
    summarise(distinct_cats = n_distinct(place_cat)) |>
    pull(distinct_cats)
  
  rows = distinct_cat
  cols = distinct_cat
  if (distinct_cat == 4) {
    cats <- c("1-5", "6-10", "11-15", "16-20")
  } else if (distinct_cat == 5) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25")
  } else if (distinct_cat == 6) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")
  } else if (distinct_cat == 7) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
              "31-35")
  } else if (distinct_cat == 8) {
    cats <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30",
              "31-35", "36-40")
  }
  
  pass_matrix <- matrix(0, nrow = rows, ncol = cols)
  for (i in 1:rows) {
    for (j in 1:cols) {
      matrix_row <- placecat |>
        filter(place_cat == cats[i] & place_next_cat == cats[j])
      if (nrow(matrix_row) == 0) {
        matrix_value <- 0
      } else {
        matrix_value <- placecat |> 
          filter(place_cat == cats[i] & place_next_cat == cats[j]) |> pull(prob)
      }
      pass_matrix[i, j] <- matrix_value
    }
  }
  
  rownames(pass_matrix) <- cats
  colnames(pass_matrix) <- cats
  
  pass_matrix_df <- as.data.frame(as.table(pass_matrix))
  ggplot(pass_matrix_df, aes(Var1, Var2, fill = Freq)) +
    geom_tile() +
    scale_fill_gradient(low = "white", high = "#da3a32") +
    geom_text(aes(label = round(Freq,2))) +
    labs(title = "Passing Probabilities",
         subtitle = input_rider, 
         x = "From Place", 
         y = "To Place", fill = "Probability")
}

pass_prob_rider_fn("TANNER WARD")
```

**Q2**: What is the probability of going from one place group to another over the course of x laps?

We can let x be any positive integer. As an example, we will consider the probabilities over the course of 5 laps.

```{r}
#| output: false
x = 5
```

<u>450 class</u>:

```{r}
round(matrix450_2 %^% x, 2)
```

<u>250 class</u>:

```{r}
round(matrix250_2 %^% x, 2)
```

<u>WMX class</u>:

```{r}
round(matrixwmx_2 %^% x, 2)
```

**Q3**: Do passing probabilities themselves change over the course of a moto?

To narrow the scope of this problem, we will consider how the probability of staying in 1-5 position group changes throughout a moto.

```{r}
#| warning: false
all_2024_placecat_laps <- all_2024_complete |>
  drop_na(place) |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place_next) |>
  select(place, place_next, lap, class) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                     ifelse(place >= 6 & place <= 10, "6-10",
                     ifelse(place >= 11 & place <= 15, "11-15",
                     ifelse(place >= 16 & place <= 20, "16-20",
                     ifelse(place >= 21 & place <= 25, "21-25",
                     ifelse(place >= 26 & place <= 30, "26-30",
                     ifelse(place >= 31 & place <= 35, "31-35",
                            "36-40"))))))),
         place_next_cat = ifelse(place_next >= 1 & place_next <= 5, "1-5",
                          ifelse(place_next >= 6 & place_next <= 10, "6-10",
                          ifelse(place_next >= 11 & place_next <= 15, "11-15",
                          ifelse(place_next >= 16 & place_next <= 20, "16-20",
                          ifelse(place_next >= 21 & place_next <= 25, "21-25", 
                          ifelse(place_next >= 26 & place_next <= 30, "26-30",
                          ifelse(place_next >= 31 & place_next <= 35, "31-35", "36-40")))))))) |>
  select(place_cat, place_next_cat, lap, class) |>
  group_by(place_cat, place_next_cat, lap, class) |>
  summarise(n = n()) |>
  ungroup()

all_2024_placecat_laps2 <- all_2024_placecat_laps |>
  group_by(class, lap, place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_placecat_laps3 <- all_2024_placecat_laps |>
  left_join(all_2024_placecat_laps2) |>
  mutate(prob = n/total) |>
  filter(place_cat == "1-5" & place_next_cat == "1-5")

all_2024_placecat_laps3 |>
  ggplot(aes(x = lap, y = prob, color = class)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Lap", y = "Probability of Retaining 1-5", color = "Class")
```

The bold line is the line of best fit from local polynomial regression. A big takeaway from this visual is that it usually becomes less likely to change positions the longer the race goes on. This makes sense as gaps begin to open up.

**Q4**: What is the probability of starting in place X and finishing the moto in place Y?

We define the starting place as the place the rider is in after the first lap of the moto.

<u>450 class</u>:

```{r}
#| warning: false
all_2024_450_start_finish <- all_2024_complete |>
  filter(class == "450" & lap == 1) |>
  select(place, finish)

all_2024_450_start_finish_sum <- all_2024_450_start_finish |>
  group_by(place, finish) |>
  summarise(n = n()) |>
  ungroup()

all_2024_450_start_finish_sum_2 <- all_2024_450_start_finish_sum |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_450_start_finish_sum_3 <- all_2024_450_start_finish_sum |>
  left_join(all_2024_450_start_finish_sum_2) |>
  mutate(prob = n/total)

rows = 30
cols = 30
cats <- 1:30

matrix450_3 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2024_450_start_finish_sum_3 |> 
      filter(place == i & finish == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_450_start_finish_sum_3 |> 
        filter(place == i & finish == j) |> pull(prob)
    }
    matrix450_3[i, j] <- matrix_value
  }
}

rownames(matrix450_3) <- cats
colnames(matrix450_3) <- cats
print(round(matrix450_3, 2))

all_2024_450_start_finish_cat <- all_2024_complete |>
  filter(class == "450" & lap == 1) |>
  select(place, finish) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                           ifelse(place >= 6 & place <= 10, "6-10",
                           ifelse(place >= 11 & place <= 15, "11-15",
                           ifelse(place >= 16 & place <= 20, "16-20",
                           ifelse(place >= 21 & place <= 25, 
                                  "21-25","26-30"))))),
         finish_cat = ifelse(finish >= 1 & finish <= 5, "1-5",
                      ifelse(finish >= 6 & finish <= 10, "6-10",
                      ifelse(finish >= 11 & finish <= 15, "11-15",
                      ifelse(finish >= 16 & finish <= 20, "16-20",
                      ifelse(finish >= 21 & finish <= 25, "21-25", 
                             "26-30")))))) |>
  select(place_cat, finish_cat) |>
  group_by(place_cat, finish_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2024_450_start_finish_cat_2 <- all_2024_450_start_finish_cat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_450_start_finish_cat_3 <- all_2024_450_start_finish_cat |>
  left_join(all_2024_450_start_finish_cat_2) |>
  mutate(prob = n/total)

rows2 = 6
cols2 = 6
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")

matrix450_4 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2024_450_start_finish_cat_3 |>
      filter(place_cat == cats2[i] & finish_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_450_start_finish_cat_3 |> 
        filter(place_cat == cats2[i] & finish_cat == cats2[j]) |> pull(prob)
    }
    matrix450_4[i, j] <- matrix_value
  }
}

rownames(matrix450_4) <- cats2
colnames(matrix450_4) <- cats2

sf_450 <- as.data.frame(as.table(matrix450_4))
ggplot(sf_450, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Start-Finish Probabilities",
       subtitle = "450", x = "Start ", y = "Finish",
       fill = "Probability")
```

<u>250 class</u>:

```{r}
#| warning: false
all_2024_250_start_finish <- all_2024_complete |>
  filter(class == "250" & lap == 1) |>
  select(place, finish)

all_2024_250_start_finish_sum <- all_2024_250_start_finish |>
  group_by(place, finish) |>
  summarise(n = n()) |>
  ungroup()

all_2024_250_start_finish_sum_2 <- all_2024_250_start_finish_sum |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_250_start_finish_sum_3 <- all_2024_250_start_finish_sum |>
  left_join(all_2024_250_start_finish_sum_2) |>
  mutate(prob = n/total)

rows = 40
cols = 40
cats <- 1:40

matrix250_3 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2024_250_start_finish_sum_3 |> 
      filter(place == i & finish == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_250_start_finish_sum_3 |> 
        filter(place == i & finish == j) |> pull(prob)
    }
    matrix250_3[i, j] <- matrix_value
  }
}

rownames(matrix250_3) <- cats
colnames(matrix250_3) <- cats
print(round(matrix250_3, 2))

all_2024_250_start_finish_cat <- all_2024_complete |>
  filter(class == "250" & lap == 1) |>
  select(place, finish) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                           ifelse(place >= 6 & place <= 10, "6-10",
                           ifelse(place >= 11 & place <= 15, "11-15",
                           ifelse(place >= 16 & place <= 20, "16-20",
                           ifelse(place >= 21 & place <= 25, "21-25",
                           ifelse(place >= 26 & place <= 30, "26-30",
                           ifelse(place >= 31 & place <= 35, "31-35", 
                                  "36-40"))))))),
         finish_cat = ifelse(finish >= 1 & finish <= 5, "1-5",
                      ifelse(finish >= 6 & finish <= 10, "6-10",
                      ifelse(finish >= 11 & finish <= 15, "11-15",
                      ifelse(finish >= 16 & finish <= 20, "16-20",
                      ifelse(finish >= 21 & finish <= 25, "21-25", 
                      ifelse(finish >= 26 & finish <= 30, "26-30",
                      ifelse(finish >= 31 & finish <= 35, "31-35", 
                             "36-40")))))))) |>
  select(place_cat, finish_cat) |>
  group_by(place_cat, finish_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2024_250_start_finish_cat_2 <- all_2024_250_start_finish_cat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_250_start_finish_cat_3 <- all_2024_250_start_finish_cat |>
  left_join(all_2024_250_start_finish_cat_2) |>
  mutate(prob = n/total)

rows2 = 8
cols2 = 8
cats2 <- c("1-5", "6-10", "11-15", "16-20", 
           "21-25", "26-30", "31-35", "36-40")

matrix250_4 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2024_250_start_finish_cat_3 |>
      filter(place_cat == cats2[i] & finish_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_250_start_finish_cat_3 |> 
        filter(place_cat == cats2[i] & finish_cat == cats2[j]) |> pull(prob)
    }
    matrix250_4[i, j] <- matrix_value
  }
}

rownames(matrix250_4) <- cats2
colnames(matrix250_4) <- cats2

sf_250 <- as.data.frame(as.table(matrix250_4))
ggplot(sf_250, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Start-Finish Probabilities",
       subtitle = "250", x = "Start ", y = "Finish",
       fill = "Probability")
```

<u>WMX class</u>:

```{r}
#| warning: false
all_2024_wmx_start_finish <- all_2024_complete |>
  filter(class == "WMX" & lap == 1) |>
  select(place, finish)

all_2024_wmx_start_finish_sum <- all_2024_wmx_start_finish |>
  group_by(place, finish) |>
  summarise(n = n()) |>
  ungroup()

all_2024_wmx_start_finish_sum_2 <- all_2024_wmx_start_finish_sum |>
  group_by(place) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_wmx_start_finish_sum_3 <- all_2024_wmx_start_finish_sum |>
  left_join(all_2024_wmx_start_finish_sum_2) |>
  mutate(prob = n/total)

rows = 30
cols = 30
cats <- 1:30

matrixwmx_3 <- matrix(0, nrow = rows, ncol = cols)
for (i in 1:rows) {
  for (j in 1:cols) {
    matrix_row <- all_2024_wmx_start_finish_sum_3 |> 
      filter(place == i & finish == j)
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_wmx_start_finish_sum_3 |> 
        filter(place == i & finish == j) |> pull(prob)
    }
    matrixwmx_3[i, j] <- matrix_value
  }
}

rownames(matrixwmx_3) <- cats
colnames(matrixwmx_3) <- cats
print(round(matrixwmx_3, 2))

all_2024_wmx_start_finish_cat <- all_2024_complete |>
  filter(class == "WMX" & lap == 1) |>
  select(place, finish) |>
  mutate(place_cat = ifelse(place >= 1 & place <= 5, "1-5",
                           ifelse(place >= 6 & place <= 10, "6-10",
                           ifelse(place >= 11 & place <= 15, "11-15",
                           ifelse(place >= 16 & place <= 20, "16-20",
                           ifelse(place >= 21 & place <= 25, 
                                  "21-25","26-30"))))),
         finish_cat = ifelse(finish >= 1 & finish <= 5, "1-5",
                      ifelse(finish >= 6 & finish <= 10, "6-10",
                      ifelse(finish >= 11 & finish <= 15, "11-15",
                      ifelse(finish >= 16 & finish <= 20, "16-20",
                      ifelse(finish >= 21 & finish <= 25, "21-25", 
                             "26-30")))))) |>
  select(place_cat, finish_cat) |>
  group_by(place_cat, finish_cat) |>
  summarise(n = n()) |>
  ungroup()

all_2024_wmx_start_finish_cat_2 <- all_2024_wmx_start_finish_cat |>
  group_by(place_cat) |>
  summarise(total = sum(n)) |>
  ungroup()

all_2024_wmx_start_finish_cat_3 <- all_2024_wmx_start_finish_cat |>
  left_join(all_2024_wmx_start_finish_cat_2) |>
  mutate(prob = n/total)

rows2 = 6
cols2 = 6
cats2 <- c("1-5", "6-10", "11-15", "16-20", "21-25", "26-30")

matrixwmx_4 <- matrix(0, nrow = rows2, ncol = cols2)
for (i in 1:rows2) {
  for (j in 1:cols2) {
    matrix_row <- all_2024_wmx_start_finish_cat_3 |>
      filter(place_cat == cats2[i] & finish_cat == cats2[j])
    if (nrow(matrix_row) == 0) {
      matrix_value <- 0
    } else {
      matrix_value <- all_2024_wmx_start_finish_cat_3 |> 
        filter(place_cat == cats2[i] & finish_cat == cats2[j]) |> pull(prob)
    }
    matrixwmx_4[i, j] <- matrix_value
  }
}

rownames(matrixwmx_4) <- cats2
colnames(matrixwmx_4) <- cats2

sf_wmx <- as.data.frame(as.table(matrixwmx_4))
ggplot(sf_wmx, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#da3a32") +
  geom_text(aes(label = round(Freq,2))) +
  labs(title = "Start-Finish Probabilities",
       subtitle = "WMX", x = "Start ", y = "Finish",
       fill = "Probability")
```

## 4.3 Rider fatique

Do riders get more tired as the moto goes on? We can check this by checking whether lap times increase with lap number?

We filter out the first lap since it is not a full lap of the moto.

<u>450 class</u>:

We also filter out lap times over 200 seconds as these are outliers and might affect the model.

```{r}
#| warning: false
all_2024_complete |>
  filter(class == "450" & lap > 1) |>
  ggplot(aes(x = lap, y = time, 
             color = paste0(round, "-", moto))) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(x = "Lap", y = "Lap Time", color = "Race ID",
       title = "All 450 Lap Times by Race")

all_2024_complete |>
  filter(class == "450" & time < 200 & lap > 1) |>
  group_by(paste0(round, "-", moto)) |>
  do(model = lm(time ~ lap, data = .)) |>
  mutate(slope = coef(model)["lap"]) |>
  select(`paste0(round, "-", moto)`, slope) |>
  mutate(across(slope, round, 2)) |>
  gt() |>
  cols_label(`paste0(round, "-", moto)` = "Race ID",
             slope = "Slope")
```

This shows that over the course of motos, lap times of riders are likely to increase. This is an indication of fatique. Note that the moto 2-2 ended early due to Dylan Wright's fall so rider fatigue was unlikely to kick in.

Here are the average lap times by lap for each moto:

```{r}
#| warning: false
all_2024_complete |>
  filter(class == "450" & time < 200 & lap > 1) |>
  group_by(paste0(round, "-", moto), lap) |>
  summarize(avg_time = mean(time)) |>
  ungroup() |>
  ggplot(aes(x = lap, y = avg_time, 
             color = `paste0(round, "-", moto)`)) +
  geom_line() +
  labs(x = "Lap", y = "Average Lap Time", color = "Race ID",
       title = "450 - Average Lap Times")
```

This shows that average lap times are increasing up until the last couple of laps of the moto. I intuit that this is for a couple of reasons. First, we need to consider survivorship bias. Because lapping other riders exists, the final couple of lap times are only going to include the top riders in that moto, making the average lap time much faster here. Second, at the end of a moto, riders are giving it everything they can as they know the race is ending soon.

<u>250 class</u>:

We also filter out lap times over 250 seconds as these are outliers and might affect the model.

```{r}
#| warning: false
all_2024_complete |>
  filter(class == "250" & lap > 1) |>
  ggplot(aes(x = lap, y = time, 
             color = paste0(round, "-", moto))) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(x = "Lap", y = "Lap Time", color = "Race ID",
       title = "All 450 Lap Times by Race")

all_2024_complete |>
  filter(class == "250" & time < 250 & lap > 1) |>
  group_by(paste0(round, "-", moto)) |>
  do(model = lm(time ~ lap, data = .)) |>
  mutate(slope = coef(model)["lap"]) |>
  select(`paste0(round, "-", moto)`, slope) |>
  mutate(across(slope, round, 2)) |>
  gt() |>
  cols_label(`paste0(round, "-", moto)` = "Race ID",
             slope = "Slope")
```

This shows that for most motos, lap times of riders are likely to increase in time. This is an indication of fatigue.

Here are the average lap times by lap for each moto:

```{r}
#| warning: false
all_2024_complete |>
  filter(class == "250" & time < 250 & lap > 1) |>
  group_by(paste0(round, "-", moto), lap) |>
  summarize(avg_time = mean(time)) |>
  ungroup() |>
  ggplot(aes(x = lap, y = avg_time, 
             color = `paste0(round, "-", moto)`)) +
  geom_line() +
  labs(x = "Lap", y = "Average Lap Time", color = "Race ID",
       title = "250 - Average Lap Times")
```

This shows that average lap times are increasing up until the last couple of laps of the moto, as in the 450 class.

<u>WMX class</u>:

We also filter out lap times over 250 seconds as these are outliers and might affect the model.

```{r}
#| warning: false
all_2024_complete |>
  filter(class == "WMX" & lap > 1) |>
  ggplot(aes(x = lap, y = time, 
             color = paste0(round, "-", moto))) +
  geom_point(alpha = 0.2) +
  geom_smooth() +
  labs(x = "Lap", y = "Lap Time", color = "Race ID",
       title = "All 450 Lap Times by Race")

all_2024_complete |>
  filter(class == "WMX" & time < 250 & lap > 1) |>
  group_by(paste0(round, "-", moto)) |>
  do(model = lm(time ~ lap, data = .)) |>
  mutate(slope = coef(model)["lap"]) |>
  select(`paste0(round, "-", moto)`, slope) |>
  mutate(across(slope, round, 2)) |>
  gt() |>
  cols_label(`paste0(round, "-", moto)` = "Race ID",
             slope = "Slope")
```

This shows that for most motos, lap times of riders are likely to decrease in time. Unlike the men's classes, this is an indication that riders are not fatiguing since they are getting faster.

Here are the average lap times by lap for each moto:

```{r}
#| warning: false
all_2024_complete |>
  filter(class == "WMX" & time < 250 & lap > 1) |>
  group_by(paste0(round, "-", moto), lap) |>
  summarize(avg_time = mean(time)) |>
  ungroup() |>
  ggplot(aes(x = lap, y = avg_time, 
             color = `paste0(round, "-", moto)`)) +
  geom_line() +
  labs(x = "Lap", y = "Average Lap Time", color = "Race ID",
       title = "WMX - Average Lap Times")
```

## 4.4 Rider rivalries

Which two riders in each class battled the most throughout the season? In other words, add up the number of laps where the behind times between adjacent pairs of riders is less than 2 seconds. Which pair spend the most laps in this state?

<u>450 class</u>:

```{r}
#| warning: false
gaps_450_1 <- all_2024_complete |>
  filter(class == "450") |>
  drop_na(place) |>
  select(lap, time, behind, place, race_id, rider) |>
  group_by(race_id, rider) |>
  mutate(total_time = cumsum(time)) |>
  ungroup() |>
  mutate(next_place = place + 1)

gaps_450_2 <- gaps_450_1 |>
  left_join(gaps_450_1, by = join_by(next_place == place, 
                                     race_id == race_id,
                                     lap == lap)) |>
  select(lap, race_id, rider.x, total_time.x, place, 
         rider.y, total_time.y, next_place) |>
  mutate(diff = total_time.x - total_time.y) |>
  drop_na(rider.y) |>
  filter(diff >= -2)

rival_pairs_450 <- gaps_450_2 |> 
  select(rider.x, rider.y) |>
  mutate(pair = map2_chr(rider.x, rider.y, 
                         ~paste(sort(c(.x, .y)), collapse = "-"))) |>
  count(pair) |>
  separate(pair, into = c("rider1", "rider2"), sep = "-") |>
  arrange(desc(n))

rival_pairs_450 |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(rider1 = "Rider A",
             rider2 = "Rider B",
             n = "# of Laps")
  
```

<u>250 class</u>:

```{r}
#| warning: false
gaps_250_1 <- all_2024_complete |>
  filter(class == "250") |>
  drop_na(place) |>
  select(lap, time, behind, place, race_id, rider) |>
  group_by(race_id, rider) |>
  mutate(total_time = cumsum(time)) |>
  ungroup() |>
  mutate(next_place = place + 1)

gaps_250_2 <- gaps_250_1 |>
  left_join(gaps_250_1, by = join_by(next_place == place, 
                                     race_id == race_id,
                                     lap == lap)) |>
  select(lap, race_id, rider.x, total_time.x, place, 
         rider.y, total_time.y, next_place) |>
  mutate(diff = total_time.x - total_time.y) |>
  drop_na(rider.y) |>
  filter(diff >= -2)

rival_pairs_250 <- gaps_250_2 |> 
  select(rider.x, rider.y) |>
  mutate(pair = map2_chr(rider.x, rider.y, 
                         ~paste(sort(c(.x, .y)), collapse = "-"))) |>
  count(pair) |>
  separate(pair, into = c("rider1", "rider2"), sep = "-") |>
  arrange(desc(n))

rival_pairs_250 |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(rider1 = "Rider A",
             rider2 = "Rider B",
             n = "# of Laps")
  
```

<u>WMX class</u>:

```{r}
#| warning: false
gaps_wmx_1 <- all_2024_complete |>
  filter(class == "WMX") |>
  drop_na(place) |>
  select(lap, time, behind, place, race_id, rider) |>
  group_by(race_id, rider) |>
  mutate(total_time = cumsum(time)) |>
  ungroup() |>
  mutate(next_place = place + 1)

gaps_wmx_2 <- gaps_wmx_1 |>
  left_join(gaps_wmx_1, by = join_by(next_place == place, 
                                     race_id == race_id,
                                     lap == lap)) |>
  select(lap, race_id, rider.x, total_time.x, place, 
         rider.y, total_time.y, next_place) |>
  mutate(diff = total_time.x - total_time.y) |>
  drop_na(rider.y) |>
  filter(diff >= -2)

rival_pairs_wmx <- gaps_wmx_2 |> 
  select(rider.x, rider.y) |>
  mutate(pair = map2_chr(rider.x, rider.y, 
                         ~paste(sort(c(.x, .y)), collapse = "-"))) |>
  count(pair) |>
  separate(pair, into = c("rider1", "rider2"), sep = "-") |>
  arrange(desc(n))

rival_pairs_wmx |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(rider1 = "Rider A",
             rider2 = "Rider B",
             n = "# of Laps")
  
```

## 4.5 Points race

```{r}
#| warning: false
standings_450 <- tibble(rider = c("JESS PETTIS", "JOSIAH NATZKE", 
                                 "PRESTON KILROY","AARON TANTI", 
                                 "TANNER WARD", "PARKER EALES",
                                 "DYLAN WRIGHT", "JULIEN BENEK", 
                                 "DANIEL ELMORE", "MITCHELL HARRISON"),
                 total_points = c(355, 344, 300, 252, 234, 
                                  189, 180, 176, 176, 117))

points_race_450 <- all_2024_complete |>
  filter(class == "450") |>
  distinct(race_id, rider, points) |>
  filter(rider %in% standings_450$rider)

for (i in standings_450$rider) {
  for(j in points_race_450$race_id) {
    check <- points_race_450 |> filter(race_id == j & rider == i)
    if(nrow(check) == 0) {
            points_race_450 <- points_race_450 |> add_row(race_id = j,
                                                          rider = i,
                                                          points = 0)
    }
  }
}

points_race_450 |>
  mutate(season_moto = if_else(race_id == "450_2024_1", 1,
                       if_else(race_id == "450_2024_2", 2,
                       if_else(race_id == "450_2024_3", 3,
                       if_else(race_id == "450_2024_4", 4,
                       if_else(race_id == "450_2024_5", 5,
                       if_else(race_id == "450_2024_6", 6,
                       if_else(race_id == "450_2024_7", 7,
                       if_else(race_id == "450_2024_8", 8,
                       if_else(race_id == "450_2024_9", 9,
                       if_else(race_id == "450_2024_10", 10,
                       if_else(race_id == "450_2024_11", 11,
                       if_else(race_id == "450_2024_12", 12,
                       if_else(race_id == "450_2024_13", 13,
                       if_else(race_id == "450_2024_14", 14,
                       if_else(race_id == "450_2024_15", 
                               15,16)))))))))))))))) |>
  arrange(season_moto, rider) |>
  group_by(rider) |>
  mutate(total_points = cumsum(points)) |>
  ungroup() |>
  ggplot(aes(x = season_moto, y = total_points, color = rider)) +
  geom_line() +
  labs(x = "Moto #", y = "Total Points", title = "450 Points Race",
       color = "Rider")

```

```{r}
#| warning: false
standings_250 <- tibble(rider = c("KAVEN BENOIT", "DREW ADAMS", 
                                 "GAGE LINVILLE", "SEBASTIEN RACINE",
                                 "WYATT KERR", "QUINN AMYOTTE",
                                 "TYLER MEDAGLIA", "TIGER WOOD", 
                                 "DYLAN REMPEL", "EVAN STICE"),
                 total_points = c(351, 349, 254, 225, 223, 
                                  215, 178, 162, 155, 141))

points_race_250 <- all_2024_complete |>
  filter(class == "250") |>
  distinct(race_id, rider, points) |>
  filter(rider %in% standings_250$rider)

for (i in standings_250$rider) {
  for(j in points_race_250$race_id) {
    check <- points_race_250 |> filter(race_id == j & rider == i)
    if(nrow(check) == 0) {
            points_race_250 <- points_race_250 |> add_row(race_id = j,
                                                          rider = i,
                                                          points = 0)
    }
  }
}

points_race_250 |>
  mutate(season_moto = if_else(race_id == "250_2024_1", 1,
                       if_else(race_id == "250_2024_2", 2,
                       if_else(race_id == "250_2024_3", 3,
                       if_else(race_id == "250_2024_4", 4,
                       if_else(race_id == "250_2024_5", 5,
                       if_else(race_id == "250_2024_6", 6,
                       if_else(race_id == "250_2024_7", 7,
                       if_else(race_id == "250_2024_8", 8,
                       if_else(race_id == "250_2024_9", 9,
                       if_else(race_id == "250_2024_10", 10,
                       if_else(race_id == "250_2024_11", 11,
                       if_else(race_id == "250_2024_12", 12,
                       if_else(race_id == "250_2024_13", 13,
                       if_else(race_id == "250_2024_14", 14,
                       if_else(race_id == "250_2024_15", 
                               15,16)))))))))))))))) |>
  arrange(season_moto, rider) |>
  group_by(rider) |>
  mutate(total_points = cumsum(points)) |>
  ungroup() |>
  ggplot(aes(x = season_moto, y = total_points, color = rider)) +
  geom_line() +
  labs(x = "Moto #", y = "Total Points", title = "250 Points Race",
       color = "Rider")

```

The WMX class was not included in this analysis since there were technically three different series - West, East, and E/W Shootout.

## 4.6 Positions gained

```{r}
#| warning: false
all_2024_finish <- all_2024_complete |>
  filter(! race_id %in% c("wmx_2024_41", "wmx_2024_42")) |>
  filter(lap == 1) |>
  select(rider, finish, place, race_id, class) |>
  group_by(rider, class) |>
  summarize(avg_finish = round(mean(finish), 2)) |>
  ungroup()

all_2024_poschange <- all_2024_complete |>
  filter(! race_id %in% c("wmx_2024_41", "wmx_2024_42")) |>
  mutate(race_id_name = paste(race_id, rider, sep = "_")) |>
  mutate(place_next = lead(place, n = 1L,
                           order_by = race_id_name)) |>
  group_by(race_id_name) |>
  mutate(place_next = c(place_next[-n()], NA)) |>
  ungroup() |>
  drop_na(place, place_next) |>
  mutate(places_gain = if_else(place_next < place,
                               place - place_next, 0),
         places_lost = if_else(place_next > place,
                               place_next - place, 0)) |>
  group_by(rider, class) |>
  summarize(total_motos = n_distinct(race_id),
            total_passes = sum(places_gain),
            passes_per_moto = round(total_passes/total_motos,2)) |>
  arrange(desc(passes_per_moto)) |>
  ungroup() |>
  left_join(all_2024_finish)
```

<u>450 class</u>:

```{r}
#| warning: false

all_2024_poschange |>
  filter(class == "450") |>
  select(rider, total_motos, total_passes,
         avg_finish, passes_per_moto) |>
  arrange(desc(avg_finish)) |>
  slice_min(avg_finish, n = 15) |>
  gt() |>
  cols_label(rider = "Rider",
             total_motos = "Total Motos",
             total_passes = "Total Passes",
             avg_finish = "Average Finish",
             passes_per_moto = "Passes/Moto")
```

<u>250 class</u>:

```{r}
#| warning: false

all_2024_poschange |>
  filter(class == "250") |>
  select(rider, total_motos, total_passes,
         avg_finish, passes_per_moto) |>
  arrange(desc(avg_finish)) |>
  slice_min(avg_finish, n = 15) |>
  gt() |>
  cols_label(rider = "Rider",
             total_motos = "Total Motos",
             total_passes = "Total Passes",
             avg_finish = "Average Finish",
             passes_per_moto = "Passes/Moto")
```

<u>WMX class</u>:

```{r}
#| warning: false

all_2024_poschange |>
  filter(class == "WMX") |>
  select(rider, total_motos, total_passes,
         avg_finish, passes_per_moto) |>
  arrange(desc(avg_finish)) |>
  slice_min(avg_finish, n = 15) |>
  gt() |>
  cols_label(rider = "Rider",
             total_motos = "Total Motos",
             total_passes = "Total Passes",
             avg_finish = "Average Finish",
             passes_per_moto = "Passes/Moto")
```

```{r}
#| warning: false
ggplot(all_2024_poschange,
       aes(x = avg_finish,
           y = passes_per_moto,
           color = class)) +
  geom_point() +
  geom_smooth(method = "loess") +
  labs(x = "Average Finish",
       y = "Passes per Moto",
       color = "Class")
```

There is lots of useful insight from this graph. First, we see a similar pattern for all three classes. The riders who finish near the front are not making many passes. They usually have good starts and stay in the front throughout the moto. The riders who finish near the rear are not making many passes. They usually have poor starts and stay in the back throughout the moto. The most passing occurs in the middle of pack.

Second, we see that the 250 class has the most passing and the WMX class has the least amount of passing.

## 4.7 Best starters

As before, since we don't access to any holeshot data, we define the start variable as the rider's place after the first lap.

```{r}
#| warning: false
all_2024_startfinish <- all_2024_complete |>
  filter(! race_id %in% c("wmx_2024_41", "wmx_2024_42")) |>
  filter(lap == 1) |>
  select(rider, finish, place, race_id, class) |>
  group_by(rider, class) |>
  summarize(total_motos = n_distinct(race_id),
            avg_start = round(mean(place), 2),
            avg_finish = round(mean(finish), 2)) |>
  ungroup()
```

<u>450 class</u>:

```{r}
#| warning: false

all_2024_startfinish |>
  filter(class == "450") |>
  select(rider, total_motos, avg_start,
         avg_finish) |>
  arrange(desc(avg_start)) |>
  slice_min(avg_start, n = 15) |>
  gt() |>
  cols_label(rider = "Rider",
             total_motos = "Total Motos",
             avg_start = "Average Start",
             avg_finish = "Average Finish")
```

<u>250 class</u>:

```{r}
#| warning: false

all_2024_startfinish |>
  filter(class == "250") |>
  select(rider, total_motos, avg_start,
         avg_finish) |>
  arrange(desc(avg_start)) |>
  slice_min(avg_start, n = 15) |>
  gt() |>
  cols_label(rider = "Rider",
             total_motos = "Total Motos",
             avg_start = "Average Start",
             avg_finish = "Average Finish")
```

<u>WMX class</u>:

```{r}
#| warning: false

all_2024_startfinish |>
  filter(class == "WMX") |>
  select(rider, total_motos, avg_start,
         avg_finish) |>
  arrange(desc(avg_start)) |>
  slice_min(avg_start, n = 15) |>
  gt() |>
  cols_label(rider = "Rider",
             total_motos = "Total Motos",
             avg_start = "Average Start",
             avg_finish = "Average Finish")
```

We can see that those who start strong often finish strong. For the 250 and 450 class, the best starter was also the series champion.

We can also see in which direction the rider usually heads after the start. If the average finish is higher than the average start, the rider usually drops back. If the average finish is lower than the average start, the rider usually climbs up. However, it is important to note that a rider who starts well usually has no place to go but down. This is why Jess Pettis has an average finish below his average start because he starts most motos in the lead.

## 4.8 Lap consistency

By creating density histograms of lap times during a particular moto, we can assess which riders are consistently fast.

Here are the top five finishers of the first moto for the 450 class:

```{r}
#| warning: false

moto_consistency <- function(input_raceid) {
  all_2024_complete |>
    filter(race_id == input_raceid) |>
    filter(lap != 1 & finish <= 5) |>
    ggplot(aes(x = time, color = paste0(finish, "-", rider))) +
    geom_density() +
    labs(x = "Lap Time",
         y = "Density",
         color = "Rider",
         title = "450 class - Rd 1 - Moto 1")
}
moto_consistency("450_2024_1")
```

The central tendency of the race winner, Dylan Wright, is the further to the left, indicating that he had the fastest lap times. However, Wright also has the flattest curve, indicating a lack of consistency in this moto. The curve of Josiah Natzke is the tallest, which shows that he had the most consistent lap times, even if they weren't the fastest in the field.

## 4.9 Lap comparisons to race leader

It is useful to compare how a rider is performing in a moto relative to the race leader. We can see the points in a race where the rider begins to fall off.

The following table highlights 450 rider Tanner Ward's lap times in the first moto of the season.

```{r}
#| warning: false

lap_comparison <- function(input_raceid, input_rider) {
  leader_times <- all_2024_complete |>
    filter(race_id == input_raceid & place == 1) |>
    select(lap, time) |>
    rename(leader_time = time) |>
    arrange(lap)
  
  all_2024_complete |>
    filter(race_id == input_raceid & rider == input_rider) |>
    select(lap, time) |>
    left_join(leader_times) |>
    mutate(last_lap_time = lag(time, default = NA)) |>
    mutate(diff_lastlap = time - last_lap_time,
           diff_leader = time - leader_time) |>
    select(lap, time, diff_lastlap, diff_leader) |>
    gt() |>
    cols_label(lap = "Lap",
             time = "Time",
             diff_lastlap = "Difference to Last Lap",
             diff_leader = "Difference to Leader Lap")
}

lap_comparison("450_2024_1", "TANNER WARD")
```

We can see that Ward had a slower start to this moto, putting in laps around 4-5 seconds behind the leader. But he comes on strong towards the end, even putting in some laps faster than the race leader Dylan Wright.

## 4.10 Win probability model

If we know how many laps are left, what place a rider is positioned in, and how far they are behind the leader, we can predict the likelihood that they go on to win the race.

```{r}
#| warning: false

winning_model_data <- all_2024_complete |>
  filter(! is.na(behind)) |>
  mutate(won_race = if_else(finish == 1, 1, 0)) |>
  group_by(race_id, rider) |>
  mutate(laps_left = max(lap) - lap) |>
  ungroup() |>
  select(rider, race_id, laps_left, place, behind, won_race) |>
  mutate(won_race = factor(won_race))

logistic_model <- glm(won_race ~ laps_left + place + behind, 
                      data = winning_model_data, 
                      family = binomial)
summary(logistic_model)

exp(coef(logistic_model))
```

Because the p-value is less than 0.05, we can say that all included variables are statistically significant.

Some interpretation:

- For each additional lap left, the odds of winning decrease by 7%. Riders are more likely to win when there are fewer laps left. 
- For each increase in place number, the odds of winning decrease by 30%. Being in a higher place number lowers the chance of winning the moto. 
- For each additional second behind the leader, the odds of winning decrease by 27%. Being further behind the leader reduces the chances of winning.

Now, we can add win probability predictions to the data and see which riders average the highest win probability.

```{r}
#| warning: false

winning_model_data$win_prob <- round(predict(logistic_model, 
                                       type = "response"), 2)

winning_model_data |>
  group_by(rider, race_id) |>
  summarize(avg_win_prob = mean(win_prob)) |>
  ungroup() |>
  group_by(rider) |>
  summarize(avg_win_prob = round(mean(avg_win_prob),2)) |>
  slice_max(avg_win_prob, n = 15) |>
  gt() |>
  cols_label(rider = "Rider",
             avg_win_prob = "Average Win Probability")
```

Here are the cases where the rider has the largest range in win probability.

```{r}
winning_model_data |>
  group_by(rider, race_id) |>
  mutate(low_prob = min(win_prob),
         high_prob = max(win_prob),
         prob_diff = high_prob - low_prob) |>
  ungroup() |>
  distinct(rider, race_id, low_prob, high_prob, prob_diff) |>
  slice_max(prob_diff, n = 15) |>
  gt() |>
  cols_label(rider = "Rider",
             race_id = "Race ID",
             low_prob = "Lowest Win Prob",
             high_prob = "Highest Win Prob",
             prob_diff = "Win Prob Difference")
```

## 4.11 Rider profiles

<u>450 class</u>:

```{r}
#| warning: false
moto_id_450 <- tibble(race_id = c("450_2024_1", "450_2024_2",
                                  "450_2024_3", "450_2024_4",
                                  "450_2024_5", "450_2024_6",
                                  "450_2024_7", "450_2024_8",
                                  "450_2024_9", "450_2024_10",
                                  "450_2024_11", "450_2024_12",
                                  "450_2024_13", "450_2024_14",
                                  "450_2024_15", "450_2024_16"),
                      moto_id = c("11","12","21","22","31","32","41","42",
                                  "51","52","61","62","71","72","81","82"))

profile <- function(rider_name, num, mx_class, joiner) {
  profile_1 <- all_2024_complete |>
    filter(class == mx_class & lap == 1) |>
    select(race_id, rider, location, place, finish, points) |>
    filter(rider == rider_name) |>
    full_join(joiner) |>
    arrange(moto_id) |>
    select(rider, location, moto_id, place, finish, points) |>
    pivot_longer(cols = c("place", "finish"), 
                 names_to = "category", 
                 values_to = "value") |>
    mutate(category = ifelse(category == "place", "Start",
                             "Finish"))
  
  profile_1$category <- factor(profile_1$category, 
                               levels = c("Start", "Finish"))
  
  riders_class <- all_2024_complete |>
    filter(class == mx_class) |>
    distinct(rider) |>
    cross_join(joiner)
  
  profile_2 <- all_2024_complete |>
    filter(class == mx_class & lap == 1) |>
    select(race_id, rider, place, finish, points) |>
    full_join(riders_class) |>
    group_by(rider) |>
    mutate(total_points = cumsum(points)) |>
    ungroup() |>
    group_by(race_id) |>
    mutate(points_rank = rank(desc(total_points), 
                              ties.method = "first")) |>
    filter(rider == rider_name) |>
    arrange(moto_id) |>
    select(rider, moto_id, points_rank)
  
  local <- profile_1 |>
    distinct(location) |>
    pull(1)
  
  gaps_1 <- all_2024_complete |>
    filter(class == mx_class) |>
    drop_na(place) |>
    select(lap, time, behind, place, race_id, rider) |>
    group_by(race_id, rider) |>
    mutate(total_time = cumsum(time)) |>
    ungroup() |>
    mutate(next_place = place + 1)
  
  gaps_2 <- gaps_1 |>
    left_join(gaps_1, by = join_by(next_place == place, 
                                   race_id == race_id,
                                   lap == lap)) |>
    select(lap, race_id, rider.x, total_time.x, place, 
           rider.y, total_time.y, next_place) |>
    mutate(diff = total_time.x - total_time.y) |>
    drop_na(rider.y) |>
    filter(diff >= -3)
  
  rival_pairs <- gaps_2 |> 
    select(rider.x, rider.y) |>
    mutate(pair = map2_chr(rider.x, rider.y, 
                           ~paste(sort(c(.x, .y)), collapse = "-"))) |>
    count(pair) |>
    separate(pair, into = c("rider1", "rider2"), sep = "-") |>
    arrange(desc(n)) |>
    filter(rider1 == rider_name | rider2 == rider_name) |>
    mutate(rival = if_else(rider1 == rider_name,
                           rider2, rider1)) |>
    distinct(rival) |>
    pull(rival)
  
  ggplot() +
    geom_bar(data = profile_1, 
             aes(x = factor(moto_id), y = value, fill = category),
             stat = "identity", position = "dodge") +
    geom_line(data = profile_2, 
              aes(x = factor(moto_id), y = points_rank, 
                  group = 1, color = "Standings"), 
              size = 0.8) +
    geom_text(data = profile_1,
              aes(x = factor(moto_id), y = 0, fill = category,
                  label = value, color = category),
              position = position_dodge(width = 0.9), 
              vjust = -0.5, size = 2, color = "black") +
    labs(x = "Moto ID", y = "Position", fill = " ", color = " ",
         title = paste0("#", num, " ", rider_name),
         subtitle = paste0("Class: ", mx_class, "      ",
                           "Location: ", local)) +
    scale_fill_manual(values = c("Start" = "#32D2DA", 
                                 "Finish" = "#da3a32")) +
    scale_color_manual(values = c("Standings" = "black")) +
    scale_y_continuous(name = "Position", 
                       breaks = seq(0, max(max(profile_1$value,
                                               profile_2$points_rank, na.rm = TRUE)),
                                    by = 1),
                       sec.axis = sec_axis(trans = ~ ., 
                                           name = " ", 
                                           breaks = seq(0, max(max(profile_2$points_rank, profile_1$value, na.rm = TRUE)), by = 1))) +
    theme_minimal() +
    geom_label(aes(x = Inf, y = Inf, 
                   label = paste0("Top Rivalries:\n1. ", rival_pairs[1], "\n2. ", rival_pairs[2], "\n3. ", rival_pairs[3])), 
               fill = "white",
               vjust = 1, hjust = 1, size = 2)
  
}

profile("JESS PETTIS", "15", "450", moto_id_450)
profile("TANNER WARD", "84", "450", moto_id_450)
```

<u>250 class</u>:

```{r}
#| warning: false

moto_id_250 <- tibble(race_id = c("250_2024_1", "250_2024_2",
                                  "250_2024_3", "250_2024_4",
                                  "250_2024_5", "250_2024_6",
                                  "250_2024_7", "250_2024_8",
                                  "250_2024_9", "250_2024_10",
                                  "250_2024_11", "250_2024_12",
                                  "250_2024_13", "250_2024_14",
                                  "250_2024_15", "250_2024_16"),
                      moto_id = c("11","12","21","22","31","32","41","42",
                                  "51","52","61","62","71","72","81","82"))

profile("DREW ADAMS", "300", "250", moto_id_250)
profile("TYLER MEDAGLIA", "5", "250", moto_id_250)
```

# 5 Extensions

There are so many opportunities for further analysis. Here are some ideas I have to expand this project in the future:

-   Add previous years data to the analysis. According to the [results](https://triplecrownseries.ca/results/) tab of the Triple Crown Series website, there is data going back to 2018.

-   Integrate weather and temperature data to explore correlations with rider performance, especially fatigue levels.

-   More modelling such as passing and finish place prediction.

# 6 Notes

-   In the WMX class, lap times and places for Round 4 at Gopher Dunes were not available. Only finish positions were included.

# 7 Acknowledgements

Thank you to Dr. Jack Davis for suggesting the idea of transition matrices. He recommended me to his own analysis on the game, [Snakes and Ladders](https://www.stats-et-al.com/2017/11/snakes-and-ladders-and-transition.html).

Thank you [Stats F1](https://www.statsf1.com/) for rider profile ideas, such as the grid and finish position graph.
